Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-09-01 paper.Rnw
working on preparing paper for submission

% Paper header (title and authors)
\title{Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations}
\author[1]{Katherine Goode*}
\author[1,2]{Heike Hofmann}
\authormark{Goode and Hofmann}
\address[1]{\orgdiv{Department of Statistics}, \orgname{Iowa State University}, \orgaddress{\state{Iowa}, \country{United States}}}
\address[2]{\orgdiv{Center for Statistics and Applications in Forensic Evidence (CSAFE)}, \orgname{Iowa State University}, \orgaddress{\state{Iowa}, \country{United States}}}
\corres{*Corresponding author. \email{kgoode@iastate.edu}}

% Abstract and keywords
\abstract[Summary]{The importance of providing explanations for predictions made by black-box models has led to the development of model explainer methods such as LIME (local interpretable model-agnostic explanations) \citep{ribeiro:2016}. LIME uses a surrogate model to explain the relationship between predictor variables and predictions from a black-box model in a local region around a prediction of interest. However, the quality of the resulting explanations relies on how well the explainer model captures the black-box model in a specified local region. Here we introduce three visual diagnostics to assess the quality of LIME explanations: (1) explanation scatterplots, (2) assessment metric plots, and (3) feature heatmaps. We apply the visual diagnostics to a forensics bullet matching dataset to show examples where LIME explanations depend on the tuning parameter values and the explainer model oversimplifies the black-box model. Our examples raise concerns about claims made of LIME \citep{ribeiro:2016} that are similar to other criticisms in the literature \citep{alvarezmelis:2018, laugel:2018, molnar:2019}.}
\keywords{LIME, black-box models, interpretability, diagnostics}

@@

# Specify global options
knitr::opts_chunk$set(
  echo = FALSE,
  fig.keep = 'high',
  message = FALSE,
  dev = "tiff",
  fig.align = 'center'
)

# Load packages
library(caret)
library(cowplot)
library(dplyr)
library(forcats)
library(ggforce)
library(ggplot2)
library(ggpcp) #devtools::install_github("yaweige/ggpcp", build_vignettes = TRUE)
library(glmnet)
library(gower)
library(graphics)
library(gretchenalbrecht) #devtools::install_github("dicook/gretchenalbrecht")
library(latex2exp)
library(lime) #devtools::install_github("goodekat/lime")
library(limeaid) #devtools::install_github("goodekat/limeaid")
library(purrr)
library(randomForest)
library(stringr)
library(tidyr)

@@

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(data = lime_data_gathered,
             mapping = aes(x = feature_stnd_value,
                           y = prediction,
                           size = distance_value,
                           alpha = distance_value,
                           color = distance)) +
  geom_abline(data = coefs_data %>% filter(case == "good"),
              mapping = aes(intercept = int, slope = slope),
              size = 1) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c(NA, "grey30"), guide = "none") +
  geom_point(data = prediction_of_interest_gathered,
             mapping = aes(x = feature_value,
                           y = prediction,
                           fill = color),
             color = "black",
             size = 5,
             shape = 23,
             alpha = 0.75) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(feature_stnd_value = 1.2,
                      prediction = 1.6,
                      feature = c("Feature 1", "Feature 2"),
                      slope = c(paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "good",
                                               feature == "Feature 1") %>%
                                        pull(slope) %>%
                                        round(3)),
                                paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "good",
                                               feature == "Feature 2") %>%
                                        pull(slope) %>%
                                        round(3)))),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = 4) +
  labs(x = "",
       y = "Black-Box Prediction",
       title = "Conceptual Depiction of a Faithful Local Explainer Model",
       subtitle = paste("Gower Distance Metric Exponent:", good_gower_power),
       fill = "",
       alpha = "Weight",
       size = "Weight") +
  theme_linedraw(base_family = "Times", base_size = 10) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text = element_text(color = 'black', size = 10)) +
  guides(fill = guide_legend(order = 1),
         size = guide_legend(order = 2),
         alpha = guide_legend(order = 2))

@@

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(data = lime_data_gathered,
             mapping = aes(x = feature_stnd_value,
                           y = prediction,
                           size = distance_value,
                           alpha = distance_value,
                           color = distance)) +
  geom_abline(data = coefs_data %>% filter(case == "bad"),
              mapping = aes(intercept = int, slope = slope),
              size = 1) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c("grey30", NA), guide = "none") +
  geom_point(data = prediction_of_interest_gathered,
             mapping = aes(x = feature_value,
                           y = prediction,
                           fill = color),
             color = "black",
             size = 5,
             shape = 23,
             alpha = 0.75) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(feature_stnd_value = 1.2,
                      prediction = 1.6,
                      feature = c("Feature 1", "Feature 2"),
                      slope = c(paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "bad",
                                               feature == "Feature 1") %>%
                                        pull(slope) %>%
                                        round(3)),
                                paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "bad",
                                               feature == "Feature 2") %>%
                                        pull(slope) %>%
                                        round(3)))),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = 4) +
  labs(x = "",
       y = "Black-Box Prediction",
       title = "Conceptual Depiction of an Unfaithful Local Explainer Model",
       subtitle = paste("Gower Distance Metric Exponent:", bad_gower_power),
       fill = "",
       alpha = "Weight",
       size = "Weight") +
  theme_linedraw(base_family = "Times", base_size = 10) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text = element_text(color = 'black', size = 10)) +
  guides(fill = guide_legend(order = 1),
         size = guide_legend(order = 2),
         alpha = guide_legend(order = 2))

@@

# Plot the training data observed classes
sine_plot_obs <- ggplot(sine_data_train, aes(x = x1, y = x2, color = y)) +
  geom_point(alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black") +
  scale_colour_manual(values = c(sinecolor_blue, sinecolor_red)) +
  theme_bw(base_family = "Times", base_size = 10) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = TeX("y"),
       title = "Training Data")

# Plot the testing data rf predictions
sine_plot_pred <- ggplot() +
  geom_point(data = sine_poi %>%
               mutate(shape = "Prediction \nof Interest"),
             mapping = aes(x = x1,
                           y = x2,
                           #fill = rfprob_blue,
                           shape = shape),
             size = 5,
             alpha = 0.8,
             color = "black") +
  geom_point(data = sine_data_test %>% filter(y != rfpred),
             mapping = aes(x = x1, y = x2),
             shape = 1,
             size = 3,
             color = "black") +
  geom_point(data = sine_data_test,
             mapping =  aes(x = x1,
                            y = x2,
                            color = rfprob_blue,
                            fill = rfprob_blue),
             shape = 21,
             alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black") +
  scale_color_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_fill_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_shape_manual(values = 23) +
  theme_bw(base_family = "Times", base_size = 10) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = "Random \nForest \nProbability \nfor 'blue'",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       title = "Testing Data",
       shape = "") +
  guides(shape = guide_legend(order = 1))

@@

sine_poi_exp <- sine_poi_lime_default %>% plot_features() + theme(text = element_text(family = "Times", size = 10))

exp_scatter <- 
  plot_explain_scatter(sine_poi_lime_default, alpha = 0.4) +
  facet_grid(switch = "both") +
  labs(x = TeX("x_1"),
       y = TeX("x_2"),
       size = "Weight",
       shape = "",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       color = "Random \nForest \nProbability \nfor 'blue'",
       linetype = "",
       title = "Explanation Scatterplot") +
  theme_bw(base_family = "Times", base_size = 10) +
  guides(shape = guide_legend(order = 1),
         size = guide_legend(nrow = 2, byrow = T)) +
  theme(aspect.ratio = 1)

@@

<<sine-feat-heat, out.width = '3.125in', fig.width = 4, fig.height = 4.5>>=
plot_feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background = element_rect(color = "white", fill = "white")) +
  labs(y = "Case")
@
\caption{\emph{Feature Heatmap}. An example feature heatmap of the explanations from the applications of LIME with different tuning parameter values to the \data \ test set. The cases from the test set are plotted on the y-axis, and the tuning parameter values (simulation methods here) are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and tuning parameter value. The plot is faceted by the first and second most important features selected by LIME (top and bottom facets, respectively). The vertical facets separate the  kernel density method from the quantile bin methods. The vertical striping indicates that the LIME explanations are not consistent across tuning parameter values.}
\label{fig:sine-feat-heat}
\end{figure}

@@

\begin{figure*}[!thp]
<<bullet-eois, out.width = '6.5in', fig.width = 12, fig.height = 17, out.width="90%", message = FALSE>>=
# Join the plots
plot_grid(
  pf2_3qb,
  pf2_3eb,
  eoi2_3qb,
  eoi2_3eb,
  pf1_3qb,
  pf1_3eb,
  eoi1_3qb,
  eoi1_3eb,
  nrow = 4,
  rel_heights = c(0.2, 0.3, 0.2, 0.3)
)
@
\caption{Plots of LIME explanations (first and third rows) and explanation scatterplots (second and fourth rows) for  cases M and NM in the bullet test data for two tuning parameter values: 3 quantile bins (first column) and 3 equal bins (second column).  The plots provide insights into the LIME explanations and allow for the assessment of the explanation quality.}
\label{fig:bullet-eois}
\end{figure*}

>>>>

% Paper header (title and authors)
\title{Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations}
\author[1]{Katherine Goode*}
\author[1,2]{Heike Hofmann}
\authormark{Goode and Hofmann}
\address[1]{\orgdiv{Department of Statistics}, \orgname{Iowa State University}, \orgaddress{\state{Iowa}, \country{United States}}}
\address[2]{\orgdiv{Center for Statistics and Applications in Forensic Evidence (CSAFE)}, \orgname{Iowa State University}, \orgaddress{\state{Iowa}, \country{United States}}}
\corres{*Katherine Goode, Department of Statistics, Iowa State University, Ames, IA. \email{kgoode@iastate.edu}}

% Abstract and keywords
\abstract[Abstract]{The importance of providing explanations for predictions made by black-box models has led to the development of model explainer methods such as LIME (local interpretable model-agnostic explanations). LIME uses a surrogate model to explain the relationship between predictor variables and predictions from a black-box model in a local region around a prediction of interest. However, the quality of the resulting explanations relies on how well the explainer model captures the black-box model in a specified local region. Here we introduce three visual diagnostics to assess the quality of LIME explanations: (1) explanation scatterplots, (2) assessment metric plots, and (3) feature heatmaps. We apply the visual diagnostics to a forensics bullet matching dataset to show examples where LIME explanations depend on the tuning parameter values and the explainer model oversimplifies the black-box model. Our examples raise concerns about claims made of LIME that are similar to other criticisms in the literature.}
\keywords{LIME, black-box models, interpretability, diagnostics}

@@

# Specify global options
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.keep = 'high',
  fig.align = 'center',
  dev = c("cairo_pdf", "cairo_ps"),
  dev.args = list(fallback_resolution = 800),
  dpi = 800
)

# Load packages
library(Cairo)
library(caret)
library(cowplot)
library(dplyr)
library(forcats)
library(ggforce)
library(ggplot2)
library(ggpcp) #devtools::install_github("yaweige/ggpcp", build_vignettes = TRUE)
library(glmnet)
library(gower)
library(graphics)
library(gretchenalbrecht) #devtools::install_github("dicook/gretchenalbrecht")
library(latex2exp)
library(lime) #devtools::install_github("goodekat/lime")
library(limeaid) #devtools::install_github("goodekat/limeaid")
library(purrr)
library(randomForest)
library(stringr)
library(tidyr)

@@

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(data = lime_data_gathered,
             mapping = aes(x = feature_stnd_value,
                           y = prediction,
                           size = distance_value,
                           alpha = distance_value,
                           color = distance)) +
  geom_abline(data = coefs_data %>% filter(case == "good"),
              mapping = aes(intercept = int, slope = slope),
              size = 1) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c(NA, "grey30"), guide = "none") +
  geom_point(data = prediction_of_interest_gathered,
             mapping = aes(x = feature_value,
                           y = prediction,
                           fill = color),
             color = "black",
             size = 5,
             shape = 23,
             alpha = 0.75) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(feature_stnd_value = 1.2,
                      prediction = 1.6,
                      feature = c("Feature 1", "Feature 2"),
                      slope = c(paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "good",
                                               feature == "Feature 1") %>%
                                        pull(slope) %>%
                                        round(3)),
                                paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "good",
                                               feature == "Feature 2") %>%
                                        pull(slope) %>%
                                        round(3)))),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = 4) +
  labs(x = "",
       y = "Black-Box Prediction",
       title = "Conceptual Depiction of a Faithful Local Explainer Model",
       subtitle = paste("Gower Distance Metric Exponent:", good_gower_power),
       fill = "",
       alpha = "Weight",
       size = "Weight") +
  theme_linedraw(base_family = "Times", base_size = 8) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text = element_text(color = 'black', size = 10)) +
  guides(fill = guide_legend(order = 1),
         size = guide_legend(order = 2),
         alpha = guide_legend(order = 2))

@@

# Plot of good explainer model
ggplot() +
  facet_grid(. ~ feature, switch = "x") +
  geom_point(data = lime_data_gathered,
             mapping = aes(x = feature_stnd_value,
                           y = prediction,
                           size = distance_value,
                           alpha = distance_value,
                           color = distance)) +
  geom_abline(data = coefs_data %>% filter(case == "bad"),
              mapping = aes(intercept = int, slope = slope),
              size = 1) +
  scale_alpha(range = c(0.3, 1)) +
  scale_color_manual(values = c("grey30", NA), guide = "none") +
  geom_point(data = prediction_of_interest_gathered,
             mapping = aes(x = feature_value,
                           y = prediction,
                           fill = color),
             color = "black",
             size = 5,
             shape = 23,
             alpha = 0.75) +
  scale_fill_manual(values = "#FAAA72") +
  geom_text(
    data = data.frame(feature_stnd_value = 1.2,
                      prediction = 1.6,
                      feature = c("Feature 1", "Feature 2"),
                      slope = c(paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "bad",
                                               feature == "Feature 1") %>%
                                        pull(slope) %>%
                                        round(3)),
                                paste("Slope:",
                                      coefs_data %>%
                                        filter(case == "bad",
                                               feature == "Feature 2") %>%
                                        pull(slope) %>%
                                        round(3)))),
    mapping = aes(x = feature_stnd_value,
                  y = prediction,
                  label = slope),
    family = "Times",
    size = 4) +
  labs(x = "",
       y = "Black-Box Prediction",
       title = "Conceptual Depiction of an Unfaithful Local Explainer Model",
       subtitle = paste("Gower Distance Metric Exponent:", bad_gower_power),
       fill = "",
       alpha = "Weight",
       size = "Weight") +
  theme_linedraw(base_family = "Times", base_size = 8) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.placement = "outside",
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text = element_text(color = 'black', size = 10)) +
  guides(fill = guide_legend(order = 1),
         size = guide_legend(order = 2),
         alpha = guide_legend(order = 2))

@@

# Plot the training data observed classes
sine_plot_obs <- ggplot(sine_data_train, aes(x = x1, y = x2, color = y)) +
  geom_point(alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black") +
  scale_colour_manual(values = c(sinecolor_blue, sinecolor_red)) +
  theme_bw(base_family = "Times", base_size = 8) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = TeX("y"),
       title = "Training Data")

# Plot the testing data rf predictions
sine_plot_pred <- ggplot() +
  geom_point(data = sine_poi %>%
               mutate(shape = "Prediction \nof Interest"),
             mapping = aes(x = x1,
                           y = x2,
                           #fill = rfprob_blue,
                           shape = shape),
             size = 5,
             alpha = 0.8,
             color = "black") +
  geom_point(data = sine_data_test %>% filter(y != rfpred),
             mapping = aes(x = x1, y = x2),
             shape = 1,
             size = 3,
             color = "black") +
  geom_point(data = sine_data_test,
             mapping =  aes(x = x1,
                            y = x2,
                            color = rfprob_blue,
                            fill = rfprob_blue),
             shape = 21,
             alpha = 0.8) +
  geom_path(data = sinefun_data, aes(x = x, y = y),
            color = "black") +
  scale_color_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_fill_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5) +
  scale_shape_manual(values = 23) +
  theme_bw(base_family = "Times", base_size = 8) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       color = "Random \nForest \nProbability \nfor 'blue'",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       title = "Testing Data",
       shape = "") +
  guides(shape = guide_legend(order = 1))

@@

sine_poi_exp <- 
  sine_poi_lime_default %>% 
  plot_features() + 
  theme(text = element_text(family = "Times", size = 12))

exp_scatter <- 
  plot_explain_scatter(sine_poi_lime_default, alpha = 0.4) +
  facet_grid(switch = "both") +
  labs(x = TeX("x_1"),
       y = TeX("x_2"),
       size = "Weight",
       shape = "",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       color = "Random \nForest \nProbability \nfor 'blue'",
       linetype = "",
       title = "Explanation Scatterplot") +
  theme_bw(base_family = "Times", base_size = 12) +
  guides(shape = guide_legend(order = 1),
         size = guide_legend(nrow = 2, byrow = T)) +
  theme(aspect.ratio = 1)

@@

<<sine-feat-heat, out.width = '3.125in', fig.width = 4, fig.height = 4.5>>=
plot_feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 12) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        strip.background = element_rect(color = "white", fill = "white")) +
  labs(y = "Case")
@
\caption{\emph{Feature Heatmap}. An example feature heatmap of the explanations from the applications of LIME with different tuning parameter values to the \data \ test set. The cases from the test set are plotted on the y-axis, and the tuning parameter values (simulation methods here) are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and tuning parameter value. The plot is faceted by the first and second most important features selected by LIME (top and bottom facets, respectively). The vertical facets separate the  kernel density method from the quantile bin methods. The vertical striping indicates that the LIME explanations are not consistent across tuning parameter values.}
\label{fig:sine-feat-heat}
\end{figure}

@@

\begin{figure*}[!thp]
<<bullet-eois, out.width = '6.25in', fig.width = 12, fig.height = 17, message = FALSE>>=
# Join the plots
plot_grid(
  pf2_3qb,
  pf2_3eb,
  eoi2_3qb,
  eoi2_3eb,
  pf1_3qb,
  pf1_3eb,
  eoi1_3qb,
  eoi1_3eb,
  nrow = 4,
  rel_heights = c(0.2, 0.3, 0.2, 0.3)
)
@
\caption{Plots of LIME explanations (first and third rows) and explanation scatterplots (second and fourth rows) for  cases M and NM in the bullet test data for two tuning parameter values: 3 quantile bins (first column) and 3 equal bins (second column).  The plots provide insights into the LIME explanations and allow for the assessment of the explanation quality.}
\label{fig:bullet-eois}
\end{figure*}
